%
\section{Property Specification Language}
\label{sec:psl}
%
The formalization of the operational semantics for an abstract machine for even-loop based programs 
has several possible applications. First of all, it gives a formal meaning to complex computational models underlying widely used systems like Node.js. The informal documentation of operations like nextTick, setImmediate, etc that we found in the dev site \cite{NodeDoc} is ambiguous and difficult to parse. A transition system like the one presented in this paper gives a precise mathematical meaning to each operation in terms of evolution of configurations.
Based on this, we can use the operational semantics as a formal tool to reason about computations in the abstract machines.
For this purpose, we introduce an extension of regular expressions in order to handle values from an infinite domain, i.e., represent labels associated to values.
%
A finite state automata is a tuple $A=\tuple{Q,\Sigma,\delta,q_0}$ in which 
$Q$ is a finite set of states, $\delta:Q\times \Sigma \rightarrow Q$ is the transition relation
and $q_0$ is an initial state. 
A computation is a (possibly infinite) sequence $q_0 a_0 q_1 a_1 q_2 \ldots$ s.t. 
$q_{i+1}=\delta(q_i,a_i)$  for $i\geq 0$.
The observation of a computation $q_0 a_0 q_1 a_1 q_2 \ldots$ is the possibly infinite word 
$a_0 a_1 a_2 \ldots$ of labels occurring along the computation.
%

For an alphabet $\Sigma$, let $\widehat{\Sigma}=\{\widehat{a}|a\in\Sigma\}$.
An extended automata $EA$ is an automata $A=\tuple{Q,\Sigma\cup\widehat{\Sigma},\delta,q_0}$.
We use $\widehat{x}$ as a sort of variable.
The effect of $\widehat{x}$ is to associate a value $v$, non deterministically chosen,  
to $x$ and to instantiate $x$ with $v$ in every successive transition labeled with $x$  until the next $\widehat{x}$ label is encountered and so on.
If there are no occurrences of $\widehat{a}$ then the label is interpreted as $a$.
In other words during the unfolding of the transition relation of an extended automata 
for each label we maintain its current state. The initial state of label $a$ is $a$ itself.
However everytime we encounter $\widehat{a}$ its state is updated with some value taken from $D$.
More precisely, the configuration of an extended automata is a tuple
$\tuple{q,\rho}$ where $\rho:\Sigma\rightarrow D$.
The initial state is defined as $\tuple{q_0,id_\Sigma}$,  where $id_\Sigma(a)=a$ for each $a\in\Sigma$.
Given a configuration $\tuple{q,\rho}$, $a\in\Sigma$ and $\delta(q,a)=q'$, 
we have that $\tuple{q,\rho}\rightarrow^a\tuple{q',\rho}$.
Given a configuration $\tuple{q,\rho}$, $a\in\Sigma$ and $\delta(q,\widehat{a})=q'$, 
we have that $\tuple{q,\rho}\rightarrow^v\tuple{q',\rho'}$ where $\rho'(a)=v$ for some $v\in D$
and $\rho'(b)=\rho(b)$ for $b\neq a,b\in\Sigma$.
A computation is a sequence $\tuple{q_0,\rho_0}\alpha_0\tuple{q_1,\rho_1}\alpha_1\ldots$
such that $\tuple{q_i,\rho_i}\rightarrow^\alpha_i\tuple{q_{i+1},\rho_{i+1}}$ for $i\geq 0$.
The sequence of labels $\alpha_0\alpha_1\ldots$ is called here observation.
This simple extension of finite automata allows us to represent observations of value non-deterministically generated during a computation and passed from one callback to another. By definition, we consider only a finite number of variables in order to keep the model  as simplest as possible. 
Finally, we say that a program $P$ conforms to specification given as an extended automata $A$ iff for every computation $c$ of program $P$ in the abstract machine that generates a sequence of labels $\sigma$, there exists a computation in the extended automata with observation $\sigma$.

To validate examples extracted from Node.js code, we have written a meta-interpreter in Prolog that can explore all possible bounded executions  of the proposed model and catch unexpected execution orders with respect to specifications given in a sublanguage of the automata based language proposed in this section. The interpreter \cite{prolog} exploits the search mechanism of the Prolog runtime system in order to analyze sets of executions of a given input program. We remark that, although Node.js are at first sight sequential programs, their computation is often highly non- deterministic due to the heavy use of asynchronous operations and internal and external evants. 













